---
title: "Inferring motifs from VDJdb and building motif PWMs"
author: "M.S."
date: "12/5/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(data.table)
library(stringr)
library(stringdist)
library(reshape2)
library(igraph)
library(ggplot2)
```

## Export VDJdb to VDJtools format

Load latest DB for human

```{r}
df.vdjdb = fread("vdjdb_dump/vdjdb.slim.txt") %>%
  filter(species %in% c("HomoSapiens", "MusMusculus"),
         v.segm != "", j.segm != "", !is.na(j.segm)) %>%
  select(cdr3, v.segm, j.segm, v.end, j.start, antigen.epitope, gene, species) %>%
  unique
```

Select epitopes

```{r}
dt.epi.count = df.vdjdb %>%
  group_by(species, gene, antigen.epitope) %>%
  summarise(total = n()) %>%
  arrange(-total)

dt.epi.count %>% head(n=10)

df.vdjdb.good = df.vdjdb %>%
  merge(dt.epi.count %>% filter(total >= 30))
```
Back translate mock functions

```{r}
mock_codons <- c('GCT', 'TGT', 'GAT', 'GAA', 'TTT',
                 'GGT', 'ATT', 'CAT', 'AAA', 'TTA',
                 'ATG', 'AAT', 'CCT', 'CAA', 'CGT',
                 'TCT', 'ACT', 'GTT', 'TGG', 'TAT')

names(mock_codons) <- c('A', 'C', 'D', 'E', 'F',
                        'G', 'I', 'H', 'K', 'L',
                        'M', 'N', 'P', 'Q', 'R',
                        'S', 'T', 'V', 'W', 'Y')

mock_back_translate <- function(x) {
  paste0(mock_codons[x], collapse = "")
}

# "CASS" %>% strsplit('') %>% lapply(mock_back_translate)
```

Convert to VDJtools format

```{r}
df.vdjtools <- df.vdjdb.good %>%
  mutate(count = 1,
         cdr3aa = cdr3,
         cdr3nt = cdr3 %>% 
           strsplit('') %>% 
           lapply(mock_back_translate),
         v = v.segm, d = "", j = j.segm,
         vend = v.end, dstart = -1, dend = -1, jstart = j.start) %>% # 1-based, inclusive
  group_by(gene, antigen.epitope, species) %>%
  mutate(freq = count / sum(count)) %>%
  select(count, freq, cdr3nt, cdr3aa, v, d, j, vend, dstart, dend, jstart, 
         gene, antigen.epitope, species)
```

Write separate VDJtools file for each epitope

```{r}
system("mkdir vdjdb_export/")

df.vdjtools %>%
  group_by(gene, antigen.epitope, species) %>%
  do(data.frame(
    fwrite(., paste0("vdjdb_export/", .$species[1], ".",
                     .$gene[1], ".", .$antigen.epitope[1], ".txt"), sep = "\t", quote = F)
    ))
```

## Run TCRNET

List VDJdb split files

```{r}
filenames <- list.files("vdjdb_export/")
tmp <- str_split_fixed(filenames, fixed("."), 4)[,1:3]
df.files <- data.frame(file.name = filenames, species = tmp[,1], gene = tmp[,2]) %>%
  mutate(species = ifelse(species == "HomoSapiens", "human", "mouse"),
         gene = tolower(gene)) %>%
  mutate(input = paste0("vdjdb_export/", file.name),
         background = paste0("pools/", species, ".", gene, ".aa.txt"))
```

Run TCRNET via vdjtools
         
```{r}
df.files %>%
  group_by(background) %>%
  summarise(cmd = paste0("/software/bin/vdjtools -Xmx100G CalcDegreeStats -o 1,0,1 -g2 dummy -b ",
         background[1], " ", paste(input, collapse = " "), " tcrnet/")) %>%
  group_by(background) %>%
  do(data.frame(system(.$cmd)))
```

Load results

```{r}
df.tcrnet <- filenames %>%
  lapply(function(x) fread(paste0("tcrnet/", x)) %>%
                             mutate(file.name = x)) %>%
  rbindlist %>%
  merge(data.frame(file.name = filenames, 
                   species = tmp[,1], 
                   gene = tmp[,2],
                   antigen.epitope = tmp[,3]))
```

Adjust P-values and select enriched variants

```{r}
df.tcrnet <- df.tcrnet %>%
  group_by(species, gene, antigen.epitope) %>%
  mutate(p.adj = p.value.g) %>%
  ungroup %>%
  mutate(enriched = degree.s >= 3 & p.adj < 0.05)
```

## Analyze CDR3 graphs

Build hamming graph around enriched CDR3s

```{r}
compute_edges <- function(cdr3.1, cdr3.2, combine) {
  if (combine) {
    cdr3.1 <- c(cdr3.1, cdr3.2) %>% unique
    cdr3.2 <- cdr3.1
  }
  
  stringdistmatrix(cdr3.1, cdr3.2, 
                   method = "hamming", 
                   nthread = 80,
                   useNames = "strings") %>%
    as.matrix %>%
    melt %>% 
    filter(value == 1) %>%
    mutate(cdr3.from = as.character(Var1), 
           cdr3.to = as.character(Var2)) %>%
    select(-Var1, -Var2, -value) %>%
    unique
}

# I - between enriched and neighbors
df.graph.e <- df.tcrnet %>%
  group_by(species, gene, antigen.epitope) %>%
  do(compute_edges(.$cdr3aa[which(.$enriched)], .$cdr3aa, F)) %>%
  ungroup

# II - between all from previous graph
df.graph <- df.graph.e %>%
  group_by(species, gene, antigen.epitope) %>%
  do(compute_edges(.$cdr3.from, .$cdr3.to, T)) %>%
  ungroup
```

Select connected componenets/clusters with motifs

```{r}
compute_cc <- function(from, to) {
  cc <- data.frame(from, to) %>%
    graph_from_data_frame() %>%
    simplify %>%
    clusters
  
  data.frame(cdr3aa = names(cc$membership),
             cid = cc$membership,
             stringsAsFactors = F)
}

df.cc <- df.graph %>%
  group_by(species, gene, antigen.epitope) %>%
  do(compute_cc(.$cdr3.from, .$cdr3.to)) %>%
  ungroup %>%
  mutate(cid = paste(substr(species, 1, 1),
                     substr(gene, 3, 3),
                     antigen.epitope,
                     cid, sep = ".")) %>%
  group_by(cid) %>%
  mutate(csz = n()) %>%
  ungroup

most_freq_segm <- function(segm) {
  segm <- str_split_fixed(segm, ",", 2)[,1]
  names(sort(table(segm), decreasing=T))[1]
}

df.cc <- df.cc %>%
  merge(df.vdjdb %>% 
          mutate(cdr3aa = cdr3) %>% 
          filter(cdr3aa %in% df.cc$cdr3aa) %>%
          select(-cdr3)) %>%
  group_by(cid) %>%
  # select representative V/J
  mutate(v.segm.repr = most_freq_segm(v.segm), 
         j.segm.repr = most_freq_segm(j.segm)) %>%
  ungroup
```

Select and write large clusters

```{r}
df.cc.sel <- df.cc %>%
  filter(csz >= 5)

fwrite(df.cc.sel, "cluster_members.txt", sep = "\t")
```

## Compute position-weight matrices for motifs

Flatten CDR3 sequences

```{r}
df.flataa <- df.cc.sel %>%
  .$cdr3aa %>% 
  unique %>%
  strsplit("") %>%
  mclapply(function(x) data.frame(cdr3aa = paste0(x, collapse = ""), 
                                  aa = x, stringsAsFactors = F) %>% mutate(pos = 0:(n()-1)), mc.cores = 80) %>%
  rbindlist
```

Append sequences and compute PWMs

```{r}
df.pwms <- df.cc.sel %>%
  merge(df.flataa) %>%
  mutate(len = nchar(cdr3aa)) %>%
  group_by(species, gene, antigen.epitope,
           v.segm.repr, j.segm.repr, cid, csz, aa, pos, len) %>%
  summarise(count = n())
```

Load background PWMs for normalization

```{r}
df.bg.pwms <- expand.grid(species = unique(df.pwms$species), gene = unique(df.pwms$gene)) %>%
  mutate(species.1 = ifelse(species == "HomoSapiens", "human", "mouse"),
         gene.1 = tolower(gene)) %>%
  mutate(file.name = paste0("pools/pwms/", species.1, ".", gene.1, ".aa.aa_cdr3_pwm.txt")) %>%
  group_by(species, gene) %>%
  do(fread(.$file.name)) %>%
  mutate(v.segm.repr = v, j.segm.repr = j, count.bg = value) %>%
  select(-sample.id, -v, -j, -value) %>%
  group_by(species, gene, v.segm.repr, j.segm.repr, len) %>%
  mutate(total.bg = sum(count.bg[which(pos == 0)])) %>%
  ungroup

# for imputation, to use where we don't have data
df.bg.pwms.i <- df.bg.pwms %>%
  group_by(species, gene, len, pos, aa) %>%
  summarise(count.bg.i = sum(count.bg)) %>%
  group_by(species, gene, len) %>%
  mutate(total.bg.i = sum(count.bg.i[which(pos == 0)])) %>%
  ungroup
```

Merge with PWMs for VDJdb motifs

```{r}
df.pwms.info <- df.pwms %>%
  merge(df.bg.pwms, all.x = T) %>%
  merge(df.bg.pwms.i, all.x = T) %>%
  group_by(cid) %>%
  mutate(need.impute = sum(count.bg[which(pos == 0)], na.rm = T) == 0,
         count.bg = ifelse(is.na(count.bg), 0, count.bg),
         count.bg.i = ifelse(is.na(count.bg.i), 0, count.bg.i),
         total.bg = max(total.bg, na.rm = T),
         total.bg.i = max(total.bg.i, na.rm = T)) %>%
  ungroup
```

Compute frequencies, information and normalized information

```{r}
df.pwms.info.h <- df.pwms.info %>%
  group_by(species, antigen.epitope, gene, cid, csz,
           v.segm.repr, j.segm.repr, pos, len) %>%
  mutate(freq = count / csz, 
         freq.bg = ifelse(need.impute, (count.bg.i + 1) / (total.bg.i + 1), (count.bg + 1) / (total.bg + 1)),
         I = 1 + sum(freq * log(freq)) / log(20),
         I.norm = -sum(freq * log(freq.bg)) / log(20),
         height.I = freq * I,
         height.I.norm = freq * I.norm) %>%
  ungroup

# some plotting
df.pwms.info.h %>%
  select(species, gene, cid, len, pos, I, I.norm) %>%
  unique %>%
  melt(measure.vars = c("I", "I.norm")) %>%
  ggplot(aes(x = pos, y = value, group = paste(cid, variable),
             color = variable)) +
  geom_line(alpha = 0.5) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~len, scales = "free_x") +
  theme_bw()
```

Append MHC information

> NB present analysis doesn't consider cases when the same epitope is presented by different MHC

```{r}
data.frame(cid = df.cc.sel$cid,
           mhc.a = str_split_fixed(df.cc.sel$mhc.a, "[:,]", 2)[,1],
           mhc.b = str_split_fixed(df.cc.sel$mhc.b, "[:,]", 2)[,1],
           mhc.class = df.cc.sel$mhc.class) %>%
  group_by(cid) %>%
  summarise(mhc.class = mhc.class[1],
            mhc.a = most_freq_segm(mhc.a),
            mhc.b = most_freq_segm(mhc.b)) %>%
  merge(df.pwms.info.h) %>%
  fwrite("motif_pwms.txt", sep = "\t")
```